Sequential rank agreement methods for comparison of lists

### 2015-08-11 latex/paper.Rnw
Code update

<<fig1a, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE,eval=TRUE>>=
ysize <- 1000
nreference <- 20
## Plot the line
plot(res[1:ysize], col="black", lwd=3, type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth", cex=1.6)

@@

<<fig2a, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE, eval=FALSE>>=
ysize <- 1000
nreference <- 20
## Plot the line
plot(res2[1:ysize], col="black", lwd=3, type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

@@

<<fig2b, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE, eval=FALSE>>=
## Same as above but zoomed in
ysize <- 50
plot(res2[1:ysize], lwd=3, col="black", type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

@@

When the two assumptions are satisfied then it is clear that the
interesting part of the sequential rank agreement curves is restricted
to the depths where the percentage of censored items is
low. Generally, without additional prior knowledge about the
distribution of the censored items it seems reasonable to restrict the
attention of the sequential rank agreement to depths smaller than
$\max(d_1, \ldots, d_L)$.

Like for fully observed lists we generally expect the sequential rank
agreement to start low and then increase unless the lists are
completely unrelated (in which case the sequential rank agreement will
be constant at a high level) or if the lists mostly agree on the
ranking (in which case the sequential rank agreement will also be
constant but at a low level). For censored lists we also expect a
change-point around the depth where the lists are censored. This is an
artefact stemming from the fact that we assume that the remainder of
the lists can be replaced by a simple permutation of the missing
items.

@@

We now consider the special case where the rankings are obtained by
applying different methods to the same data set which contains one
outcome variable and a list of candidate predictors. To interpret the
sequential rank agreement values in this situation we propose two
different benchmark values corresponding to two different
hypotheses. We wish to determine if we observe better agreement than
what would be expected if there were no relevant information available
in the data.

The first reference hypothesis is
\begin{eqnarray*}
H_0  & : &  \text{The list rankings correspond to complete randomly}\\
       &  & \text{permuted lists}.
\end{eqnarray*}
This does not only assume that there is no information in data on
which the rankings are based but also that the methods are completely
independent.

Alternatively, we can remove the restriction on the independence among
the methods by only requiring that there is no information in the
ranking but that the rankings are all based on applying the
method/approaches to the same data
\begin{eqnarray*}
\widetilde H_0 & :&  \text{The list rankings are based on data containing}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}
$H_0$ is a quite unrealistic null hypothesis but we can easily obtain
realizations from that null hypothesis simply by permuting the items
within each list and then computing the sequential rank agreement for
the permuted lists. In the uncensored case each experiment contains
$L$ lists of random permutations of the numbers $1,\dots,P$. For the
censored/incomplete case we first permute the numbers $1,\dots,P$ and
then censor list $l$ at $d_l$. The sequential rank agreement curve
from the the original lists can then be compared to the pointwise rank
agreements obtained under $H_0$.

To obtain the distribution under $\widetilde H_0$ the idea is to repeat
many times the ranking procedures in uncorrelated data.  Thus, we
first permute the outcome variable of the data set on which the
rankings are based. This removes any association between the predictor
variables and the outcome. Then we apply the methods to the permuted
data set to generate $L$ new rankings and subsequently compute the
sequential rank agreement. Note that we only permute the outcomes and
preserve the structure of the candidate predictors. The randomization
approach requires that we have the original data available and as such
it may not be possible to evaluate $\widetilde H_0$ in all situations.

If the sequential rank agreement for the original data lies
substantially below the distribution of the sequential rank agreements
obtained under either $H_0$ or $\widetilde H_0$ then this suggests that
the original ranked lists agree \emph{more} than what we would expect
in data with no information.

For the analysis of the Golub data described in Section
\ref{sec:amfol}, Figure~\ref{fig:example1} shows the empirical
distributions of sequential rank agreement under $H_0$ and
$\widetilde H_0$ each based on $400$ permutations. Not surprisingly, the
sequential rank agreement under $\widetilde H_0$ is lower than the
sequential rank agreement under $H_0$ because the four methods used to
rank the data ($t$ test, logistic regression, elastic net, and MIC)
generally tend to identify (and rank high) similar predictors even if
there are only spurious associations. The two bottom panels in
Figure~\ref{fig:example1} also indicate that the observed sequential
rank agreement (the black line) is better than what would be expected
by chance for data that contain no information since it lies below the
reference areas. However, the censored data also suggests that there
may be at most 1 or 2 ranked items towards the top of the lists that
yield a result better than what would be expected (the bottom-right
plot).

@@

The average overlap is widely used in comparisons of (two) ranked
lists \citep{Fagin2003,Webber2010} and the idea behind the average
overlap closely resembles the sequential rank agreement. In this
section we present an extension of the average overlap to more that
two lists and compare this to sequential rank agreement.

The overlap among $L$ lists observed until depth $d$ is defined as the
proportion of items that are found in all $L$ lists compared to the
possible number of items found in all lists observed until depth $d$:
\begin{equation}
  O(d) = \frac{| \cap_{l=1}^L \{R_l^{-1}(r); r\leq d \} |}{d}. \label{def:overlap}
\end{equation}
Typically the overlap is only considered for pairwise comparisons but
formula \eqref{def:overlap} can be applied even when $L>2$
\citep{Bar-Ilan2006,Boulesteix2009}.

The average overlap at depth $d$ is defined as the average of the
overlaps until depth $d$,
\begin{equation}
AO(d) = \sum_{i=1}^d \frac{O(d)}{d}.
\end{equation}
This definition directly ensures that further emphasis is put on items
in the top of the lists since their item specific overlap contributes
to the calculation of average overlap at higher depths.

There are three main drawbacks of the average overlap method: First it
is highly sensitive to single items in the top of one the lists.  The
average overlap profile may change dramatically depending on the
presence or absence of similar items in the beginning of the
lists. For example, if all lists have the same item at the top then
the average overlap attains its maximum at 1, but if just one of the
lists differ then the overage overlap starts at its minimum of
0. Secondly, when the number of lists, $L$, is high then it might be
difficult to obtain a non-zero overlap (and hence a non-zero average
overlap) at the beginning of the lists because the overlap will be
zero until an item is present in all $L$ lists. Finally, the average
overlap has an interpretation in terms of moving averages of
percentages (of $L$ sets) which is somewhat less intuitive than our
proposed average rank distance.

In the following we compare the average overlap and the sequential
rank agreement using the analysis of the Golub data \citep{Golub1999}
of Section \ref{sec:amfol}. We consider a typical application of the
average overlap and thus restrict attention to only two gene rankings
obtained by the $t$ tests and logistic regression. These two marginal
modelling approaches should result in very similar gene rankings. The
first two columns of Table~\ref{tab1} show the actual top-10 ranking
of the predictors.

<<fig-ao1, echo=FALSE, fig=TRUE,include=FALSE, eval=FALSE>>=
myres <- average_overlap(inputmatrix[,1:2])
myres2 <- average_overlap(inputmatrix[-c(1,2),1:2])
xvals <- 1:500
plot(xvals, myres[xvals], xlim=c(1, 150), ylim=c(0,1), type="l", lwd=2, ylab="Average overlap", xlab="List depth")
lines(xvals[-c(1,2)], myres2[1:498], col="red", lwd=2)
@

<<fig-ao2, echo=FALSE, fig=TRUE,include=FALSE, eval=FALSE>>=
mmyres <- sra(inputmatrix[,1:2])
mmyres2 <- sra(inputmatrix[-c(1,2),1:2])
xvals <- 1:500
plot(xvals, mmyres[xvals], xlim=c(1, 150), ylim=c(0, 110), type="l", lwd=2, ylab="Sequential rank agreement", xlab="List depth")
lines(xvals[-c(1,2)], mmyres2[1:498], col="red", lwd=2)
@

>>>>

<<fig1a, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE,eval=TRUE>>=
ysize <- 1000
nreference <- 20
## Plot the line
plot(res[1:ysize], col="black", lwd=3, type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

@@

<<fig2a, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE, eval=TRUE>>=
ysize <- 1000
nreference <- 20
## Plot the line
plot(res2[1:ysize], col="black", lwd=3, type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

@@

<<fig2b, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE, eval=TRUE>>=
## Same as above but zoomed in
ysize <- 50
plot(res2[1:ysize], lwd=3, col="black", type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

@@

When the two assumptions are satisfied then it is clear that the
interesting part of the sequential rank agreement curves is restricted
to the depths where the number of censored items is
low. Generally, without additional prior knowledge about the
distribution of the censored items it seems reasonable to restrict the
attention of the sequential rank agreement to depths smaller than
$\max(d_1, \ldots, d_L)$ where at least one list provides an
uncensored item.

Like for fully observed lists we generally expect the sequential rank
agreement to start low and then increase unless the lists are
completely unrelated (in which case the sequential rank agreement will
be constant at a high level) or if the lists mostly agree on the
ranking (in which case the sequential rank agreement will also be
constant but at a low level). For partially ranked lists we also
expect a change-point around the depth where the lists are
censored. This is an artefact stemming from the fact that we assume
that the remainder of the lists can be replaced by a simple
permutation of the missing items.

@@

To interpret the sequential rank agreement values we propose two
different benchmark values corresponding to two different
hypotheses. We wish to determine if we observe better agreement than
what would be expected if there were no relevant information available
in the data.

The first reference hypothesis is
\begin{eqnarray*}
H_0  & : &  \text{The list rankings correspond to complete randomly}\\
       &  & \text{permuted lists},
\end{eqnarray*}
which not only assumes that there is no information in the data on
which the rankings are based but also that the methods are completely
independent.

Alternatively, we can remove the restriction on the independence among
the methods by only requiring that there is no information in the
ranking but that the rankings are all based on applying the
method/approaches to the same data
\begin{eqnarray*}
\widetilde H_0 & :&  \text{The list rankings are based on data containing}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}
$H_0$ is a quite unrealistic null hypothesis but we can easily obtain
realizations from that null hypothesis simply by permuting the items
within each list and then computing the sequential rank agreement for
the permuted lists. In the fully observed case each experiment
contains $L$ lists of random permutations of the items in $X$. For the
censored/partially observed case we first permute the items
$X_1,\dots,X_P$ and then censor list $l$ at $d_l$. The sequential rank
agreement curve from the the original lists can then be compared to,
say, the pointwise 95\% quantiles of the observed rank agreements
obtained under $H_0$.

To obtain the distribution under $\widetilde H_0$ the idea is to
repeat the ranking procedures for unassociated data many times.  Thus,
we first permute the outcome variable of the data set on which the
rankings are based. This removes any association between the predictor
variables and the outcome. Then we apply the methods to the permuted
data set to generate $L$ new rankings and subsequently compute the
sequential rank agreement. Note that we only permute the outcomes and
preserve the structure of the candidate predictors. The randomization
approach requires that we have the original data available and as such
it may not be possible to evaluate $\widetilde H_0$ in all situations.

If the sequential rank agreement for the original data lies
substantially below the distribution of the sequential rank agreements
obtained under either $H_0$ or $\widetilde H_0$ then this suggests that
the original ranked lists agree \emph{more} than what we would expect
in data with no information, and therefore that the information in the
lists is significantly more in agreement than what would be expected.

Figure~\ref{fig:example1} shows the empirical distributions of
sequential rank agreement under $H_0$ and $\widetilde H_0$ each based
on $400$ permutations of the Golub data described in Section
\ref{sec:amfol}. Not surprisingly, the sequential rank agreement under
$\widetilde H_0$ is lower than the sequential rank agreement under
$H_0$ because the four methods used to rank the data ($t$ test,
logistic regression, elastic net, and MIC) generally tend to identify
(and rank) similar predictors even if there are only spurious
associations. The two bottom panels in Figure~\ref{fig:example1} also
indicate that the observed sequential rank agreement (the black line)
is better than what would be expected by chance for data that contain
no information since it lies below the reference areas. However, the
censored data also suggests that there may be at most 1 or 2 ranked
items towards the top of the lists that yield a result better than
what would be expected (the bottom-right plot).

@@

The average overlap is a measure that is widely used in comparisons of
two fully observed ranked lists \citep{Fagin2003,Webber2010} and the
idea behind the average overlap closely resembles the sequential rank
agreement. In this section we present an extension of the average
overlap to more that two lists and compare this to sequential rank
agreement.

The overlap among $L$ lists observed until depth $d$ is defined as the
proportion of items that are found in all $L$ lists compared to the
possible number of items found in all lists observed until depth $d$:
\begin{equation}
  O(d) = \frac{| \cap_{l=1}^L \{R_l^{-1}(r); r\leq d \} |}{d}. \label{def:overlap}
\end{equation}
Typically the overlap is only considered for pairwise comparisons
\citep{Bar-Ilan2006,Boulesteix2009} but formula \eqref{def:overlap}
is easily extended to the situation where $L>2$ as shown above.

The average overlap at depth $d$ is defined as the average of the
overlaps until depth $d$,
\begin{equation}
AO(d) = \frac1d\sum_{i=1}^d O(d).
\end{equation}
This definition directly ensures that further emphasis is put on items
in the top of the lists since their item specific overlap contributes
to the calculation of average overlap at higher depths.

There are three main drawbacks of the average overlap method: First it
is highly sensitive to single items in the top of one the lists.  The
average overlap profile may change dramatically depending on the
presence or absence of similar items in the beginning of the
lists. For example, if all lists have the same item at the top then
the average overlap attains its maximum at 1, but if just one of the
lists differ then the overage overlap starts at its minimum of
0. Secondly, when the number of lists, $L$, is high then it might be
difficult to obtain a non-zero overlap (and hence a non-zero average
overlap) at the beginning of the lists because the overlap will be
zero until an item is present in all $L$ lists. Finally, the average
overlap has an interpretation in terms of moving averages of
percentages (of $L$ sets) which is somewhat less intuitive than our
proposed sequential rank distance.

In the following we compare the average overlap and the sequential
rank agreement using the analysis of the Golub data
\citep{Golub1999}. We consider a typical application of the average
overlap and thus restrict attention to only two gene rankings obtained
by the $t$ tests and logistic regression. These two marginal modeling
approaches should result in very similar gene rankings. The first two
columns of Table~\ref{tab1} show the actual top-10 ranking of the
predictors.

<<fig-ao1, echo=FALSE, fig=TRUE,include=FALSE, eval=TRUE>>=
myres <- average_overlap(inputmatrix[,1:2])
myres2 <- average_overlap(inputmatrix[-c(1,2),1:2])
xvals <- 1:500
plot(xvals, myres[xvals], xlim=c(1, 150), ylim=c(0,1), type="l", lwd=2, ylab="Average overlap", xlab="List depth")
lines(xvals[-c(1,2)], myres2[1:498], col="red", lwd=2)
@

<<fig-ao2, echo=FALSE, fig=TRUE,include=FALSE, eval=TRUE>>=
mmyres <- sra(inputmatrix[,1:2])
mmyres2 <- sra(inputmatrix[-c(1,2),1:2])
xvals <- 1:500
plot(xvals, mmyres[xvals], xlim=c(1, 150), ylim=c(0, 110), type="l", lwd=2, ylab="Sequential rank agreement", xlab="List depth")
lines(xvals[-c(1,2)], mmyres2[1:498], col="red", lwd=2)
@
